---
title: "U.S. COVID-19 Explorer Methodology"
output: html_document
---

[&larr; back to to covid19-explorer.org](http://covid19-explorer.org/)

## Preamble

This document contains a detailed methodology for the R package 
[*covid19.Explorer*](https://github.com/liamrevell/covid19.Explorer/) and *shiny*-based website 
[covid19-explorer.org](http://covid19-explorer.org).

In this package & web application we estimate total infections (from deaths & confirmed cases), graph the distribution
of infections over time among U.S. states, visualize COVID-19 deaths by age & by sex (as well as compared to deaths from all causes),
and analyze excess mortality through time in 2020 as a function of of age & state or jurisdiction.

Virtually all of our data come from the United States [Centers for Disease Control and Prevention](https://www.cdc.gov/) 
[National Center for Health Statistics](https://www.cdc.gov/nchs/index.htm) or the U.S. [Census Bureau](https://www.census.gov/).
Unfortunately, at this time we can only offer detailed analysis of United States COVID-19 and excess mortality data. It would
be to our great interest to compare not only U.S. states & jurisdictions, but also foreign countries; however, the wide 
variety of different ways in which these data are recorded and reported, as well as variation in data quality between countries,
make this endeavor entirely beyond the scope of our project here.

### Using this page

The purpose of this project is help users better understand the 2020 [COVID-19 pandemic](https://en.wikipedia.org/wiki/COVID-19_pandemic) 
in the United States. Users from other countries might also be interested in the project, for instance because the age distribution of 
mortality has been [broadly similar](https://doi.org/10.1038/s41598-020-73777-8) among different countries.

This project features two main types of application.

The first of these (exemplified by the *U.S. COVID-19 infections*, *State comparison*, *Plausible range*, and *Infection estimator* tabs)
consist of applications that are designed to *estimate* the true number of COVID-19 infections over the course of the pandemic. For
reasons that we elaborate below, confirmed COVID-19 cases underestimate the true number of infections, in most cases by a substantial
margin. Since there are a variety of reasons that the true number of infections, rather than simply the number of confirmed cases, would
be of interest, these applications are designed to help users apply a model to estimate the daily number of new infections, the plausible
range of new infections, the cumulative number of infections, or the daily or cumulative infections per 1M population.

Each of these applications uses a model - but it is one that must be parameterized by the user. We parameterize the model by specifying a
value or set of values for the [infection fatality ratio](https://en.wikipedia.org/wiki/Case_fatality_rate#Infection_fatality_rate) of
COVID-19, as well as an average lag time from infection to death. The default values for these model parameters are set to be fairly
reasonable, but users are encouraged to apply multiple values and examine the sensitivity of their results. In fact, this is one of the 
main purposes of the project!

The second type of application (examplified by the tabs *Deaths by age*, *Excess mortality by age*, and *By state*) do not employ
a model and exist primarily to permit the user to interact with COVID-19 death and 2020 excess mortality data, to understand the 
implications of these data, and to generate interesting or useful data visualization.

Detailed methodology for each type of analysis in this project is given in the sections below.

## Estimating infections

Since the beginning of this pandemic, it's been [widely understood](https://doi.org/10.1038/s41467-020-18272-4) that
confirmed COVID-19 cases underestimate the true number of infections, sometimes by a [very large 
margin](https://www.cdc.gov/library/covid19/092220_covidupdate.html).

This is likely due to multiple factors. One is limited testing capacity at the start of the epidemic. A second is the fact
that testing is voluntary, and many cases of SARS-CoV-2 infection present 
[asymptotically](https://doi.org/10.1038/d41586-020-03141-3) or with mild symptoms.

As such, we consider confirmed COVID-19 *deaths* to be a much more reliable indicator of disease burden than confirmed
cases; however, deaths are a *lagging indicator* of infections.

Consequently, we chose to use a model in which the number of new infections on day *i* was estimated by taking the number of 
observed COVID-19 deaths on day *i* + *k* and then dividing this quantity by the 
[infection fatality ratio](https://en.wikipedia.org/wiki/Case_fatality_rate#Infection_fatality_rate) (also called the infection fatality *rate* or IFR). In other words,
given 50 COVID-19 deaths on day *i* + *k*, and an IFR of 0.5%, we would predict that 1,000 new SARS-CoV-2 infections had occurred 
on day *i*.

Both *k*, the average lag time from infection to death (in cases of SARS-CoV-2 infections resulting in death), and IFR are to be
specified by the user. 

A reasonable lag time between infection & death is approximately three weeks. For example, during a large outbreak in 
[Melbourne, Australia](https://www.worldometers.info/coronavirus/country/australia/) the time difference between the peak 
recorded cases and peak confirmed COVID-19 deaths was ~17 days. (Infected persons normally test negative for the first few days 
following infection, so this more or less corresponds with a lag-time of three weeks.)

Likewise, IFR values ranging from about 0.3% to >1.0% have been reported over the course of the pandemic. 

For instance, [a study 
based on an early, super-spreader event in Germany](https://doi.org/10.1038/s41467-020-19509-y) estimated an IFR (corrected to 
the demographic distribution of the local population) of 0.36%. Other studies have reported higher estimated IFR. For example, 
an [Italian study](https://doi.org/10.1101/2020.04.18.20070912) estimated an overall infection fatality rate of 1.31%. In 
general, it's reasonable to suppose that IFR has [fallen through time](https://doi.org/10.1038/d41586-020-03132-4) as treatment 
of severely ill patients has improved. Likewise, even within the U.S. IFR is unlikely to be the same at a given date in different 
jurisdictions, due to differences in demographic structure between areas. We think it is reasonable to suppose an IFR no greater 
than 1.0% and declining from the start of the pandemic towards the present, with a current IFR of around 0.3%. We permit the user 
to specify a time-varying IFR by fixing the IFR at each quarter and then interpolating daily IFR between each quarter using [local 
regression smoothing](https://en.wikipedia.org/wiki/Local_regression) (LOESS).

Although this model is very simple, reporting can vary through time including regularly over the course of the week. To take
these reporting artifacts into account we used both [moving averaging](https://en.wikipedia.org/wiki/Moving_average) and 
[local regression](https://en.wikipedia.org/wiki/Local_regression) (LOESS) smoothing. Both the window for the moving average
and the LOESS smoothing parameter are controlled by the user.

The approach of using *only* confirmed COVID-19 deaths - though robust - does not permit us to estimate the true number of
infections between now - *k* days and the present. For this, we assumed a [sigmoidal](https://en.wikipedia.org/wiki/Sigmoid_function) 
relationship between time and the ratio of confirmed cases over the true number of infections. 

Since the true number of new infections cannot exceed the number of confirmed cases, logic dictates that this ratio must have a value between 0 and 1. 
We decided on a sigmoidal relationship because we assume the ratio was low early in the pandemic when confirming a new infection was 
limited primarily by testing capacity, but has probably risen (in many localities) to a more or less consistent value as test 
capacity has increased. Since getting tested is voluntary, and since many infections of SARS-CoV-2 are asymptomatic or only mildly 
symptomatic, it is unlikely to rise to near 1.0 in the U.S. regardless of testing. 

Figure 1 below shows daily estimated infections (under our model) / confirmed cases for all U.S. data over the entire course of the pandemic. Although the 
fit of the sigmoid function is relatively good, readers might note that the fitted curve *overestimates* the ratio of confirmed cases / 
infections near the start of the pandemic (i.e., to the left-hand side of the plot). This is not particularly worrisome because the most 
important part of the graph is the *right* side - since this is the part that will be use to extrapolate recent infection numbers from 
observed cases.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 1**: a) Observed U.S. daily COVID-19 deaths (red bars) and assumed IFR function (blue line). b) Ratio of confirmed COVID-19 cases / estimated infections."}
library(covid19.Explorer)
data(Data)
infection.estimator(data=Data,
	state="United States",
	ifr=c(0.015,0.01,0.007,0.006,0.006,0.006,0.006),
	las=1,cex.axis=0.8,cex.lab=0.9,plot="infection.ratio",
	window=1)
abline(h=seq(0.1,0.6,by=0.1),lty="dotted",col="grey")
```

Figure 1 seems to show a ratio of confirmed / estimated infections of about 0.45 at the present; however, the reader should keep in mind 
that in practice this value is estimated *separately* for each jurisdiction that is being analyzed, and as such might be *lower* in some
states and *higher* in others.

After fitting this sigmoidal curve to our observed and estimated cases through now - *k* days, we then turn to the last period. 
To obtain estimated infections for these days, we simply divide our observed cases from the last *k* days of data by our fitted values from the 
sigmoid curve.

Figure 2 shows the result of this analysis applied to data for the U.S. state of California.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 2**: a) Observed daily COVID-19 deaths and an assumed model of IFR in which the infection fatality ratio is initially high (~1.5%), but then declines and stabilizes at around 0.6% through the present day. b) Estimated daily infections (green), cases (blue), and deaths (red)."}
infection.estimator(data=Data,
	state="California",
	ifr=c(0.015,0.01,0.007,0.006,0.006,0.006,0.006),
	las=1,cex.axis=0.8,cex.lab=0.9)
```

In addition to computing the raw number of daily infections, this method can also be used to compute infections as a percentage of the total
population. 

For this calculation, we obtained state populations through time from the 
[U.S. Census Bureau](https://www.census.gov/data/datasets/time-series/demo/popest/2010s-state-total.html). Data is only given through 
2019, so to estimate state-level 2020 population sizes, we used a total mid-year 2020 U.S. population estimate of 
[331,002,651](https://www.worldometers.info/world-population/us-population/)) to 'correct' each 2019 state population size to a 2020 
level.

Our dataset did *not* include a separate population size estimate for Puerto Rico - so we used the values from 
[Wikipedia](https://en.wikipedia.org/wiki/Demographics_of_Puerto_Rico), and a similar correction to adjust to 2020 levels. (This 
probably resulted in a mid-2020 population size that is too *high* based on this [alternative 
source](https://ns.worldometers.info/world-population/puerto-rico-population/); however we imagine that this is a relatively small
effect.)

Finally, CDC mortality data splits New York City (NYC) from the rest of New York state. Since this contrast is [interesting](https://en.wikipedia.org/wiki/COVID-19_pandemic_in_New_York_City), we 
maintained this separation - and used a mid-2019 population estimate of ([8,336,817](https://en.wikipedia.org/wiki/New_York_City))
for NYC, and then simply assumed that the population of NYC has changed between 2015 and 2020 in proportion to the rest of the state. 
(Since they have a part:whole relationship, this seemed pretty reasonable.)

```{r, echo=FALSE}
States<-state.deaths(plot="States")
States<-States[sort(rownames(States)),]
States<-States[-which(rownames(States)=="United States"),]
knitr::kable(round(States),
	caption="**Table 1**: Estimated state population by year for 50 states, D.C., and Puerto Rico. New York is divided into two jurisdictions: New York City and New York, excluding NYC.",
	align="r", format = "html", table.attr = "style='width:75%;'",
	format.args=list(big.mark=","))
```

<br>
The question of cumulative percent infected is relevant to the (unnecessarily controversial) concept of ['herd 
immunity'](https://en.wikipedia.org/wiki/Herd_immunity). The following graph (Figure 3) shows estimated infections as a fraction of 
the total population for the U.S. state of Texas, using the same IFR model as in Figures 1 and 2.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 3**: a) Observed daily COVID-19 deaths and an assumed model of IFR. b) Estimated cumulative infections (green), cases (blue), and deaths (red), as a percentage of the total population of the state."}
infection.estimator(data=Data,
	state="Texas",
	ifr=c(0.015,0.01,0.007,0.006,0.006,0.006,0.006),
	las=1,cex.axis=0.8,cex.lab=0.9,
	cumulative=TRUE,percent=TRUE)
```

### Assumptions of the method

This model is very simple. 

We merely assume that if we *knew* the *true* number of infections and the IFR for our population of interest on day *i*, then we could
*predict* the number of deaths on day *i* + *k*, in which *k* is the lag-time from infection to death (for SARS-CoV-2 infections
leading to death). Having observed the deaths, & supposing a particular value of IFR for day *i*, we can likewise work backwards and 
*reconstruct* the most plausible number of infections for that day.

Although the model does not pre-suppose a specific value or function for IFR, it does require that one be specified by the user.
As such, it is probably worth mentioning the effect of setting an IFR value that is too *high* or too *low* compared to the true
IFR for the population of interest. 

An IFR that is *too high* (overall or at a specific time during the pandemic) will have the 
general effect of causing us to systematically *underestimate* the true number of infections. This makes sense because if we imagine
observing 50 COVID-19 deaths, an IFR of 0.5% would imply that these deaths correspond to a total of 1,000 infections. By contrast, 
a higher IFR of, say, 1.0% would only imply that only 500 infections had occurred. Assuming an IFR value that is *too low* will
(obviously) have exactly the opposite effect and thus cause us to *overestimate* the number of infections that have occurred.

We *do* assume a homogeneous value of *k* at any particular time. In fact, literature sources report lag-times between two and 
eight weeks; however, we assert that inferences by our method should not be badly off - so long as IFR does not swing about wildly 
from day to day, and so long as the number of deaths is not extremely few for any reporting period.

We likewise *do* assume a constant lag-period, *k*, through time. This assumption is perhaps a bit more dubious as it is quite 
reasonable to suspect that for a specific state or jurisdiction that as IFR falls, *k* might also increase. This is a complexity
that we explicit chose to ignore in our model.

We also assume that a more-or-less consistent fraction of COVID-19 deaths are reported as such - that is, that COVID-19 is neither
systematically under- or overreported as the cause of death at any point during the course of the pandemic. A violation of this 
assumption is not quite as grave as it might seem, however, because it can simply be 'baked in' to our model for IFR. For instance,
if we imagine that COVID-19 deaths were *under*-reported near the start of the pandemic (due to limited testing capacity), this 
can be accommodated into our model for daily infections simply by specifying a slightly lower IFR value at that time (keeping in
mind, of course, that the true IFR has generally [been falling through time](https://doi.org/10.1038/d41586-020-03132-4)).

In estimating the number of daily infections from now - *k* days to the present, we *assume* that the relationship between time (since
the first infections) and the ratio of confirmed and estimated infections is *sigmoidal* in shape (Figure 1). This is a *testable* 
assumption that seems to hold fairly well across the entire U.S. (Figure 1) and  for some jurisdictions , but less well for others. 
It is equally plausible to suspect that this ratio could shift not only as a function of time, but also as demands on testing 
capacity rise and fall with case numbers. This should be the subject of additional study, but our suspicion is that this would not 
be likely to have a large effect on our model compared to other simplifications.

## Mapping the distribution of infections across states

A hallmark feature of the U.S. COVID-19 pandemic has been the shifting geographic distribution of infections through time
among states.

To capture this dynamic, we devised a plotting method in which we overlay the daily or cumulative COVID-19 infections under our
model (outlined above), separated by state. 

We specifically selected a geographic color palette for this analysis such that 
[RGB](https://en.wikipedia.org/wiki/RGB_color_model) color values were made to vary as a function of latitude, longitude, and 
(arbitrarily) geographic distance from Florida. This is intended to have the effect of making the regional geographic
progression of infection more apparent in the graph. The results can be seen in Figure 4.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=8, units="in", dpi=150, fig.cap="**Figure 4**: Cumulative estimated infections separated by state."}
infections.by.state(data=Data,
	ifr=c(0.015,0.01,0.007,0.006,0.006,0.006,0.006),
	las=1,cex.axis=0.8,cex.lab=0.9,span=0.15)
```
	
### Assumptions of the method

This plotting method shares all the assumptions of our infection estimator, above, but adds the additional assumption that our model
of IFR is the same for all states. This assumption is quite dubious, in fact, as IFR could be expected to *rise* in locations were 
hospital resources are overtaxed by high disease burden; and, conversely, *fall* in hospitals were staff have more experience in
treating COVID-19 patients. Although we don't doubt that these nuances are important in making specific, quantitative statements
about the particular number of infections in each state, we nonetheless believe that our method is effective at capturing the 
overall geographic dynamics of the COVID-19 pandemic in the United States.

## Computing a plausible range of infection numbers

A relatively simple extension of our infection estimation method, above, is to admit *uncertainty* about the specific value of the
infection-fatality-ratio at any particular time during the pandemic, and measure the *sensitivity* of our prediction to a range of 
different values of IFR. This is valuable, because the question of the IFR for COVID-19 has been the subject of considerable
[controversy](https://doi.org/10.1136/bmj.m4509)!

This model can be design to accomodate an assumption of broad uncertainty in IFR early during the pandemic, with both decreasing
IFR, as well as falling uncertainty in IFR, towards the present. This is illustrated for data from the U.S. state of Louisiana in 
Figure 5.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 5**: a) Confirmed COVID-19 deaths. b) A corresponding plausible range of daily COVID-19 infections, under our model."}
infection.range.estimator(state="Louisiana",
	data=Data,
	ifr.low=c(0.008,0.006,0.005,0.004,0.004,0.004),
	ifr.high=c(0.022,0.014,0.009,0.007,0.007,0.007),
	span=0.15)
```

### Assumptions of this method

This plotting method shares all the assumptions of our infection estimator. 

It should be noted that although the shaded region around
the mean number of daily or cumulative infections in Figure 5 *looks* like a confidence band, it would only be valid to consider
it as such if our high and low values of the IFR through time represent a confidence interval around the true infection-fatality-ratio
(and, even then, this confidence interval would only take into account one source of uncertainty about the true daily case numbers - the
IFR).

## Comparing daily and cumulative infections between states

Another straightforward extension of our above-described model involves directly comparing daily (or cumulative) infections between states.

This could be a useful exercise because it is quite common for (particularly) [popular press 
sources](https://www.mlive.com/public-interest/2020/12/covid-19-numbers-in-michigan-and-ohio-rose-in-lockstep-this-fall-then-the-trendlines-went-in-opposite-directions.html) to attribute difference infection
dynamics in different states to one public health intervention or another. This attribution *may* be valid, but is often confounded by 
different infection dynamics through time in the different states being compared. 

The following graph (Figure 6) shows a comparison of daily deaths and estimated infections in Michigan and Ohio.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 6**: Daily confirmed COVID-19 deaths (a) or estimated infections (b) in the U.S. states of Michigan vs. Ohio."}
par(lwd=2)
compare.infections(data=Data,
	state=c("Michigan","Ohio"),
	ifr=c(0.015,0.01,0.007,0.006,0.006,0.006,0.006),
	las=1,cex.axis=0.8,cex.lab=0.9)
```

### Assumptions of this method

This plotting method shares all the assumptions of our infection estimator, and (just like our method for visualizing the geographic 
dynamics of the pandemic across *all* U.S. states) requires that we use the same IFR model for each state.

## Visualizing the number of COVID-19 deaths by age

In addition to modeling the number of infections through time, the *covid19.Explorer* [R 
package](https://github.com/liamrevell/covid19.Explorer/) and [website](http://covid19-explorer.org) also allow users to visualize the 
distribution of COVID-19 deaths by age & sex (Figure 7).

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 7**: a) Weekly confirmed COVID-19 deaths subdivided by age. b) Weekly COVID deaths (red) and deaths by other causes (blue) for all ages."}
covid.deaths(data=Data,
	plot="bar",
	las=1,cex.axis=0.8,cex.lab=0.9)
```

As well as graphing the raw number of deaths, this function can also be used to show the deaths / 1M population. To do this, we used
estimated U.S. populations by age and sex from the United States Census Bureau (Table 2).

```{r, echo=FALSE}
Ages<-Data$Age.Pop
nn<-Ages$Age.Group[Ages$Sex=="Female"]
Ages<-data.frame(Female=Ages$Total.Population[Ages$Sex=="Female"],
	Male=Ages$Total.Population[Ages$Sex=="Male"],
	Total=Ages$Total.Population[Ages$Sex=="Female"]+
	Ages$Total.Population[Ages$Sex=="Male"])
rownames(Ages)<-nn
knitr::kable(Ages,
	caption="**Table 2**: Estimated U.S. population by age group and sex, 2020.",
	align="r", format = "html", table.attr = "style='width:50%;'",
	format.args=list(big.mark=","))
```

<br>
This permits us to (for example) demonstrate the COVID-19 is not an important source of death for U.S. people under the age of 15, accounting
for only about 0.4% of all deaths (Figure 8).

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 8**: a) Cumulative confirmed COVID-19 deaths for children <15. b) Cumulative COVID deaths (red) and deaths by other causes (blue) for all ages."}
covid.deaths(age.group=c("Under 1 year","1-4 years","5-14 years"),
	data=Data,
	plot="smooth",
	show="per.capita",
	cumulative=TRUE,
	split.groups=FALSE,
	las=1,cex.axis=0.8,cex.lab=0.9)
```

By contrast, COVID-19 accounts for >10% of all deaths of U.S. persons aged &ge;75 years of age (Figure 9).

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 9**: a) Cumulative confirmed COVID-19 deaths for U.S. adults 75 years of age and older. b) Cumulative COVID deaths (red) and deaths by other causes (blue) for all ages."}
covid.deaths(age.group=c("75-84 years", "85 years and over"),
	data=Data,
	plot="smooth",
	show="per.capita",
	cumulative=TRUE,
	split.groups=FALSE,
	las=1,cex.axis=0.8,cex.lab=0.9)
```

## Excess mortality analysis

Lastly, the the *covid19.Explorer* [R package](https://github.com/liamrevell/covid19.Explorer/) and 
[website](http://covid19-explorer.org) allows users to explore excess mortality by age and state, regardless of cause.

For this analysis, we used the [2014-2018 weekly counts of deaths by state and select 
causes](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr) and the [2019-2020 provisional death 
counts](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/muzy-jte6) from the National Center for Health 
Statistics (NCHS) of the United States Centers for Disease Control and Prevention.

Provisional death counts are incomplete, particularly for recent weeks, due to the lag time between when a death occurs and when a 
death certificate is submitted to the NCHS. This accounts for the drop-off in excess deaths towards the right of each plot in 2020 data. 
This lag may also differ between different jurisdictions.

To compute the raw death counts for any jurisdiction, we simply tabulated the 2015-2018 counts (we excluded 2014) with the 2019-2020 
provisional counts. The following table gives an example of tabulated death counts for a the state of Massachusetts.

```{r, echo=FALSE}
Deaths<-state.deaths(plot="Deaths")
Deaths<-cbind(Week=1:52,Deaths)
knitr::kable(Deaths,
	caption="**Table 3**: Weekly death counts and provisional death counts in Massachusetts according to the NCHS from 2015 through 2020.",
	align="r",format = "html", table.attr = "style='width:50%;'",
	format.args=list(big.mark=","))
```

<br>
To *correct* observed deaths in prior years to 2020 levels, we simply multiplied the past-year death tally by the ratio the jurisdiction 
population in 2020 compared to the population in the past year.

Table 4 gives an example using the jurisdiction of New Jersey.

```{r, echo=FALSE}
Deaths<-state.deaths(state="New Jersey",plot="Deaths",corrected=TRUE)
Deaths<-cbind(Week=1:52,Deaths)
knitr::kable(round(Deaths,1),
	caption="**Table 4**: Weekly death counts and provisional death counts in New Jersey, corrected to New Jersey 2020 estimated population.",
	align="r",format = "html", table.attr = "style='width:50%;'",
	format.args=list(big.mark=","))
```

<br>
To compute excess deaths for any jurisdiction, we took each column of Table 3 (or Table 4, for corrected death counts) and *subtracted* 
the mean of columns 2015 through 2019. This treats 2015 through 2019 as 'normal' years, and 2020 as 
[unusual](https://en.wikipedia.org/wiki/COVID-19_pandemic). Table 5 gives an example of this calculation using the jurisdiction of Texas, and 
weekly death counts from 2015-2019 corrected to 2020 population.

```{r, echo=FALSE}
Excess<-state.deaths(state="Texas",plot="Excess",corrected=TRUE)
Excess<-cbind(Week=1:52,Excess)
knitr::kable(round(Excess,1),
	caption="**Table 5**: Weekly excess death counts and provisional death counts in Texas, compared to 2015-2019 corrected death counts.",
	align="r",format = "html", table.attr = "style='width:50%;'",
	format.args=list(big.mark=","))
```

<br>
To compute the weekly death rate per one-million population, we now divided the total number of deaths in Table 3 by the total population 
(in millions) of each jurisdiction. Note, that since we used different population sizes for each year of data (Table 2) the notion of 
'correcting' to 2020 population size is meaningless here.

Table 6 gives the weekly per 1M population death rate for the District of Columbia.

```{r, echo=FALSE}
PerCapita<-state.deaths(state="District of Columbia",plot="PerCapita")
PerCapita<-cbind(Week=1:52,PerCapita)
knitr::kable(round(PerCapita,1),
	caption="**Table 6**: Weekly deaths and provisional deaths per 1M population in the District of Columbia.",
	align="r",format = "html", table.attr = "style='width:50%;'",
	format.args=list(big.mark=","))
```
<br>
To compute the weekly excess per capita deaths (once again, per 1M population), we took the excess deaths and divided by the population 
size of the corresponding jurisdiction - in millions. The cumulative excess per capita deaths is one way of measuring the death toll of 
COVID-19.

Table 7 gives the weekly per 1M population excess death rate for New York City.

```{r, echo=FALSE}
PerCapitaExcess<-state.deaths(state="New York City",plot="PerCapitaExcess")
PerCapitaExcess<-cbind(Week=1:52,PerCapitaExcess)
knitr::kable(round(PerCapitaExcess,1),
	caption="**Table 7**: Weekly excess deaths and provisional deaths per 1M population in New York City.",
	align="r",format = "html", table.attr = "style='width:50%;'",
	format.args=list(big.mark=","))
```

<br>
Finally, to calculate the weekly percent above normal, we took the ratio of the observed death counts compare to the mean from 2015-2019. 
Once again, this treats 2015 through 2019 as 'normal' years, and 2020 as [unusual](https://en.wikipedia.org/wiki/COVID-19_pandemic). 
Table 8 gives an example of the weekly percent above normal for the jurisdiction of Louisiana. To accumulate the percent above normal, we 
cannot simply accumulate the week-by-week percentages (as we do for excess deaths). Instead we must compute the cumulative deaths in each 
year and compare them to the 2015-2019 average cumulative deaths by the same week.

```{r, echo=FALSE}
PercentAbove<-state.deaths(state="Louisiana",plot="PercentAbove")
PercentAbove<-cbind(Week=1:52,PercentAbove)
knitr::kable(round(PercentAbove,1),
	caption="**Table 8**: Weekly percent above normal deaths for Louisiana.",
	align="r",format = "html", table.attr = "style='width:40%;'",
	format.args=list(big.mark=","))
```

<br>
Figure 10 shows the weekly and cumulative percent deaths above normal for the U.S. state of Michigan.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 10**: Weekly (a) and cumulative (b) % deaths above normal for the U.S. state of Michigan."}
state.deaths(state="Michigan",data=Data,
	las=1,cex.axis=0.8,plot="percent above normal")
```

We calculated deaths above normal by *age* in exactly the same way, but in this case since we lacked detailed demographic information by
state (i.e., age-stratified population sizes), we show *only* excess deaths and percent deaths above normal. Figure 11 gives an example for
persons aged 25-44 from the U.S. state of Florida.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.width=12, fig.height=10, units="in", dpi=150, fig.cap="**Figure 11**: Weekly (a) and cumulative percent (b) deaths above normal for adults aged 25-44 in Florida during 2020."}
age.deaths(state="Florida",
	age.group="25-44 years",
	data=Data,
	las=1,cex.axis=0.8,plot="raw & percent above normal")
```

This analysis shows an approximately 30% above normal death rate in 2020 compared to 2015-2019 for adults 25-44 years of age in Florida.

### Contact & other information

The R package [*covid19.Explorer*](https://github.com/liamrevell/covid19.Explorer/) and the website [covid19-explorer.org](http://covid19-explorer.org) were developed by [Dr. Liam Revell](http://faculty.umb.edu/liam.revell/). Please [contact Liam](mailto:liam.revell@umb.edu) with questions about this page, the [covid19-explorer.org website](http://covid19-explorer.org) or the R package.

<br><br>
[&larr; back to to covid19-explorer.org](http://covid19-explorer.org/)

